Summer 2025 ToDo

### 2025-08-20 Relations/Paper/Summer_2025_TODO.tex
bugfix newline
New todo items
Progress in WF section. Work to be done.

\section*{Agda standard library}
\\ Some of the work we have explored is formalized in the Agda standard library.  
% https://agda.github.io/agda-stdlib/v1.7.3/Relation.Binary.Rewriting.html#2096
\begin{todolist}
  \item Write about the standard library formalization in our paper and talk about how ours differs. 
  \item Create an equivalences file showing that our formalization agrees with the ASL (specifically definitions such as confluet, wcr, nf, leading up to showing our formalizations of Newman's lemma agree).
  \item Investigate contributing work to the ASL. This may mean refactoring the work already done to conform with the ASL. Specific contributions would likely be: Generalized NL. Theorem's 122 and 123. Alternative definitions of Wellfounded. 
\end{todolist}

>>>>

\section*{Agda standard library}
Some of the work we have explored is formalized in the Agda standard library.  
% https://agda.github.io/agda-stdlib/v1.7.3/Relation.Binary.Rewriting.html#2096
\begin{todolist}
  \item Write about the standard library formalization in our paper and talk about how ours differs. 
  \item Create an equivalences file showing that our formalization agrees with the ASL (specifically definitions such as confluet, wcr, nf, leading up to showing our formalizations of Newman's lemma agree).
  \item Investigate contributing work to the ASL. This may mean refactoring the work already done to conform with the ASL. Specific contributions would likely be: Generalized NL. Theorem's 122 and 123. Alternative definitions of Wellfounded, the implications for the hierarchy.
\end{todolist}

\section*{20th August todos}
\begin{todolist}
  \item (Secondary) Find consistency with NP and UN with respect to reduction and conversion. Should the default be by reduction for both. Should we be explicit with $\to$ and = in each case. 
  \item Carry on modifying the names of wfXXX- to wfXXXnotnot with prefix underscore in agda and in latex.
  \item Make a decision regarding WFcor. We use it to imply WFacc using AccCor. But otherwise it is sort of orphaned off. Does it help contribute to the overall story?
\end{todolist}

### 2025-08-20 Relations/Paper/Well-foundedness.tex
Progress in WF section. Work to be done.

An abstract reduction $\to_R$ is strongly normalizing
if and only if the converse relation is well-founded.
In this section, we will compare several definitions of well-foundedness, which depend on the following properties.
\begin{definition}
  Let $R \subseteq A \times A$.  We write $Rxy$ when $(x,y) \in R$.
  \begin{enumerate}
    \item $P \subseteq A$ is \emph{inductive}
    if, for each $x \in A$, $(\forall y. Ryx \to y \in P)$ implies $x \in P$.

@@

    \item $s : \nat \to A$ is \emph{$R$-decreasing} if for all $k$, $(s(k+1),s(k)) \in R$.
  \end{enumerate}
\end{definition}
\apcomment{Add the definitions of well-foundedness here and then continue with the text as below.}
\subsection{New version: Synopsis and Outline}

@@

One could also consider proving well-foundedness ``up to $\lnot\lnot$'',
asserting that every element is $\lnot\lnot$-accessible,
every sequence $\lnot\lnot$-contains a non-decreasing index,
for every non-empty subset there $\lnot\lnot$-exists a minimal element, etc.
This cuts down some, but not all, of the complexity.
For example, the two minimality principles
described above are equivalent with this formulation.
% The implication from the inductive to sequential definition no longer requires
% decidability of $R$.
% On the other hand, the reverse
At the same time, the
implication from sequential to inductive wellfoundedness
still requires an additional hypothesis, which may be described as the
constructive contrapositive of the definition of accessibility.
Specifically, we call {\accCor} the statement that, if an element $x$
is \emph{not} accessible, then there (strongly) exists a $y$ with $Ryx$
and $y$ not accessible.  (In the context of rewriting systems,
this condition asserts the existence of a function which maps any
non-strongly normalizing element to a one-step reduct with the same property.)
Assuming \accCor, all notions of well-foundedness
are constructively equivalent up to $\lnot\lnot$.

>>>>

An abstract reduction $\to_R$ is strongly normalizing
if and only if the converse relation is well-founded.
In this section, we will compare several definitions of well-foundedness, which depend on the properties in Defintion~\ref{def:WFproperties} below.

For $P \subseteq A$, let $\ol{P} = \{ x \in A \mid x \notin P\}$ denote the complement of $P$.
\begin{definition}\label{def:WFproperties}
  Let $R \subseteq A \times A$.  We write $Rxy$ when $(x,y) \in R$.
  \begin{enumerate}
    \item $P \subseteq A$ is \emph{inductive}
    if, for each $x \in A$, $(\forall y. Ryx \to y \in P)$ implies $x \in P$.

@@

    \item $P \subseteq A$ is \emph{coreductive} if, for each $x \in \ol{P}$, there is a $y \in A$ such that $Ryx$ and $y \in \ol{P}$.

    \item $s : \nat \to A$ is \emph{$R$-decreasing} if for all $k$, $(s(k+1),s(k)) \in R$. 
  \end{enumerate}
\end{definition}
The definition of a $R$-decreasing sequence is repeated from Section~\ref{sec:Definitions}, as it is used extensively in this section.

\begin{definition}
  Let $R \subseteq A \times A$. We write $Rxy$ when $(x,y) \in R$.
  \begin{enumerate}
    \item $R$ is \emph{accessibly well-founded} ($\WFacc$) if every element is accessible:
      \[
        \forall x \in A.\; x \text{ is accessible}.
      \]
    \item $R$ is \emph{inductively well-founded} ($\WFind$) if every inductive $P \subseteq A$ is universally true:
      \[
        \forall P \subseteq A.\;\text{$P$ is inductive} \Rightarrow \forall x \in A.\; x \in P.
      \]
    \item $R$ is \emph{sequentially well-founded} ($\WFseq$) if every sequence contains an index at which it is not decreasing:
      \[
        \forall s : \nat \to A \exists k : \nat.\, \lnot \, R \,(s\,k)\,(s\,(k+1)).
      \]
    \item  $R$ is \emph{minimally well-founded} ($\WFmin$) if every nonempty subset has a $P$-minimal element:
      \[
        \forall P \subseteq A.\; P \neq \varnothing \Rightarrow
        \exists x \in P.\; \text{$x$ is $P$-minimal}.
      \]
    \item  $R$ is \emph{minimally well-founded for $\lnot \lnot$-closed predicates} ($\WFminDNE$) if every nonempty $\lnot \lnot$-closed subset contains a $P$-minimal element:
      \[
        \forall P \subseteq A.\; P \neq \varnothing \Rightarrow \left(\ol{\ol P} \subseteq P\right) \Rightarrow
        \exists x \in P.\; \text{$x$ is $P$-minimal}.
      \]
    \item $R$ is \emph{coreductively well-founded} ($\WFcor$) if every coreductive $P \subseteq A$ is universally true:
      \[
        \forall P \subseteq A.\;(\text{$P$ is coreductive}) \Rightarrow \forall x \in A.\; x \in P.
      \]

  \end{enumerate}
\end{definition}
We have formalized the above definitions of well-foundedness in \texttt{WFDefinitions.agda}.

@@

One could also consider proving well-foundedness ``up to $\lnot\lnot$'',
asserting that every element is $\lnot\lnot$-accessible,
every sequence $\lnot\lnot$-contains a non-decreasing index,
for every non-empty subset there $\lnot\lnot$-exists a minimal element, etc. These are formalized in the definition below (and in \texttt{WFWeakDefinitions.agda}).

\begin{definition}
  Let $R \subseteq A \times A$. We write $Rxy$ when $(x,y) \in R$.
  \begin{enumerate}
    \item $R$ is \emph{weakly accessibly well-founded} ($\WFaccm$) if accessibility holds up to double negation:
      \[
        \forall x \in A.\; \lnot \lnot  (x \text{ is accessible}).
      \]
    \item $R$ is \emph{weakly inductively well-founded} ($\WFindm$) if every inductive predicate holds for all elements up to double negation:
      \[
        \forall P \subseteq A.\;(\text{$P$ is inductive}) \Rightarrow \forall x \in A.\; \lnot \lnot (x \in P).
      \]
    \item $R$ is \emph{weakly sequentially well-founded} ($\WFseqm$) if no sequence is $R$-decreasing:
      \[
        \forall s:\nat \to A.\; \lnot \text{($s$ is $R$-decreasing)}.
      \]
    \item  $R$ is \emph{weakly minimally well-founded} ($\WFminm$) if every nonempty subset contains a $\lnot \lnot P$-minimal element:
      \[
        \forall P \subseteq A.\; P \neq \varnothing \Rightarrow \lnot \lnot \;
        (\exists x \in P.\; \text{$x$ is $P$-minimal}).
      \]
    \item  $R$ is \emph{weakly minimally well-founded under double negation elimination} ($\WFminDNEm$) if every nonempty $\lnot \lnot$-closed subset $\lnot \lnot$ contains a $P$-minimal element:
      \[
        \forall P \subseteq A.\; P \neq \varnothing \Rightarrow \left(\ol{\ol P} \subseteq P\right) \Rightarrow \lnot \lnot \;
        (\exists x \in P.\; \text{$x$ is $P$-minimal}).
      \]      
  \end{enumerate}
\end{definition}

\begin{figure}[ht]
\centering  
\begin{tikzpicture}[
  node distance=1cm and .5cm,
  box/.style={draw, rectangle, rounded corners, minimum width=1cm, minimum height=1cm, align=center},
  arrow/.style={-{Latex}, thick}
]

% Row 1: WFacc
\node[box] (WFacc) {WFacc};
% \node[box, right=of WFacc] (NNWFacc) {$\lnot\lnot$WFacc};
\node[box, right=2.9cm of WFacc] (WFaccM) {WFacc-};

\draw[arrow] (WFacc) -- (WFaccM);
% \draw[arrow] (NNWFacc) -- (WFaccM);
\draw[arrow, bend right] (WFaccM) to node[above, text=red] {accDNE} (WFacc);

% Row 2: WFind
% \node[box, below=of WFacc] (WFind) {WFind};
% \node[box, right=of WFind] (NNWFind) {$\lnot\lnot$WFind};
% \node[box, right=of NNWFind] (WFindM) {WFind-};

% \draw[arrow] (WFind) -- (NNWFind);
% \draw[arrow] (NNWFind) -- (WFindM);

% \draw[arrow] (NNWFseq) -- (WFseqM);

% Row 5: WFminDNE -> WFminDNE-
\node[box, below=of WFacc] (WFminDNE) {WFminDNE};
\node[box, right=2.0cm of WFminDNE] (WFminDNEM) {WFminDNE-};

\draw[arrow] (WFminDNE) -- (WFminDNEM);

% Row 4: WFmin -> WFmin-
\node[box, below=of WFminDNE] (WFmin) {WFmin};
\node[box, right=2.8cm of WFmin] (WFminM) {WFmin-};

\draw[arrow] (WFmin) -- (WFminM);

% Row 3: WFseq
\node[box, below=of WFmin] (WFseq) {WFseq};
% \node[box, right=of WFseq] (NNWFseq) {$\lnot\lnot$WFseq};
\node[box, right=2.9cm of WFseq] (WFseqM) {WFseq-};

\draw[arrow] (WFseq) -- (WFseqM);

% Row 6: WFcor 
\node[box, below=of WFseqM] (WFminCor) {WFminCor};

% Arrows between rows
% \draw[arrow, <->] (WFacc) -- (WFind);
% \draw[arrow, <->] (NNWFacc) -- (NNWFind);
% \draw[arrow, <->] (WFaccM) -- (WFindM);

\draw[arrow, bend right = 90] (WFacc) to node[left, text = red] {Rdec} (WFseq);
\draw[arrow, bend right] (WFaccM) to (WFminDNEM);
\draw[arrow, bend right = 90] (WFseqM) to node[right, text=red] {accCor} (WFaccM);  
% \draw[arrow, bend left = 55] (WFaccM) to (WFminM);

\draw[arrow] (WFmin) -- (WFseq);
\draw[arrow] (WFminM) -- (WFseqM);
\draw[arrow] (WFmin) -- (WFminDNE);
\draw[arrow, <->] (WFminM) -- (WFminDNEM);

\draw[arrow] (WFminDNE) to node[left, text = red] {accDNE} (WFacc);
\draw[arrow, bend right=45] (WFminDNE) to node[pos=0.5, left, text=red] {MP$\equiv$} (WFseq);
\draw[arrow, bend right] (WFminDNEM) to node[right, text=red] {FB} (WFaccM);
\draw[arrow, bend right] (WFseqM) to (WFminCor);
\draw[arrow, bend right] (WFminCor) to node [right, text=red] {MP$\equiv$} (WFseqM);

\end{tikzpicture}
\caption{The logical relationships between notions of well-foundedness}
 \label{fig:WF} 
\end{figure}
\clearpage

In \cite{Terese}, the $\WFseqm$ definition is used, as in the classical mathematics textbooks.

To explore how these definitions interrelate it helps to consider the following corresponding weaker definitions (which are formalized in \texttt{WFWeakDefinitions.agda}). 
The weaker definitions help us to explore how the standard well-founded definitions interrelate. This is made clear in the following diagram.

\subsection{New version: Synopsis and Outline} 
This cuts down some, but not all, of the complexity.
For example, the two minimality principles
described above are equivalent with this formulation.
% The implication from the inductive to sequential definition no longer requires
% decidability of $R$.
% On the other hand, the reverse
At the same time, the
implication from sequential to inductive wellfoundedness
still requires an additional hypothesis, which may be described as the
constructive contrapositive of the definition of accessibility.
Specifically, we call {\accCor} the statement that, if an element $x$
is \emph{not} accessible, then there (strongly) exists a $y$ with $Ryx$
and $y$ not accessible.  (In the context of rewriting systems,
this condition asserts the existence of a function which maps any
non-strongly normalizing element to a one-step reduct with the same property.)
Assuming \accCor, all notions of well-foundedness
are constructively equivalent up to $\lnot\lnot$.
